#Setup libraries for doing cluster analysis

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(factoextra)
library(cluster)
library(dplyr)
library(psych)
library(psychTools)
library(readxl)
library(knitr)
library(kableExtra)
```
#Load dataset
```{r}
df<-read_excel("loan_data.xlsx")
df_cols <- colnames(df)
set.seed(123)
```
#Load data dictionary 
```{r}
data_dict<-read_excel("Loan_data_dictionary.xlsx")
data_dict <- data_dict[data_dict$LoanStatNew %in% df_cols,]
missing_from_data_dict <- setdiff(df_cols,data_dict$LoanStatNew)
missing_from_data_dict
data_dict[nrow(data_dict) + 1,] = list(missing_from_data_dict[1],"Total revolving credit")
data_dict[nrow(data_dict) + 1,] = list(missing_from_data_dict[2],"Is the loan bad?")
new_dict_order <- as.data.frame(data_dict$LoanStatNew[order(match(data_dict$LoanStatNew,df_cols))])
colnames(new_dict_order) <- c("Variable")
colnames(data_dict) <- c("Variable", "Description")
new_data_dict <- left_join(new_dict_order,data_dict,by = "Variable")
kable(new_data_dict, format = "html", col.names = c("Variable Name", "Description"), caption= "Data Dictionary")%>%
  kable_styling(bootstrap_options = c("bordered"))
```
#Looking at the Data
```{r}
describe(df)
```
```{r}
headTail(df)
```
```{r}
#Checking Structure
str(df)
#Checking Summary
summary(df)
#Checking for NA's
as.data.frame(colSums(is.na(df)))
#Removing duplicate rows
df<- distinct(df)
```

Cleaning Data - NAs
```{r}
#counting NA's in sample
df_NAs <- as.data.frame(colSums(is.na(df)))
df_NAs$Perc_NA <- (df_NAs[,1]/nrow(df))*100
#variables with more than 0% NA
df_NAs <- df_NAs[df_NAs$Perc_NA > 0,]
#dropping columns that are more than 20% NA
drop_cols <- rownames(df_NAs[df_NAs$Perc_NA > 20,])
drop_cols
df <- df[,!(colnames(df) %in% drop_cols)]
#dropping remaining rows that contain NA values
df <- na.omit(df)
```
Removing columns that are not informative
```{r}
df$id <- NULL
df$member_id <- NULL
```
```{r}
df$grade <- as.factor(df$grade)
levels(df$grade) <- list("1"="A", "2"="B", "3" = "C", "4" = "D", "5" = "E", "6" = "F", "7" = "G")
df$grade <- as.numeric(df$grade)

```
Taking sample
```{r}
#ensuring randomness in the data 
df <- df[sample(1:nrow(df)), ]
#taking sample of 500
df_sample <- sample_n(df, 500)
#Describe sample
describe(df_sample)
```
Keep only numeric columns
```{r}
df_sample$loan_is_bad <- ifelse(df_sample$loan_is_bad == FALSE, 0,1)
df_sample_num <- df_sample %>%
  select(where(is.numeric), "loan_is_bad")
```
Looking at collinearity
```{r}
cor_matrix <- cor(df_sample_num)
df_sample_num$collections_12_mths_ex_med <- NULL
df_sample_num$policy_code <- NULL
cor_matrix2 <- cor(df_sample_num)
cor_matrix2<- round(cor_matrix2, 2)
lowerCor(df_sample_num)
```
```{r}
#Removing NA's in Corr
df_sample_num$acc_now_delinq <- NULL
lowerCor(df_sample_num)
cols_to_keep <- c("loan_amnt","term","int_rate","annual_inc","dti","delinq_2yrs","inq_last_6mths",
                  "open_acc","pub_rec","revol_bal",'revol_util', 'loan_is_bad')
df_sample_num <- df_sample_num %>% select(all_of(cols_to_keep))
lowerCor(df_sample_num)
```
Outlier detection  using mahalanobis distance 
```{r}
df_sample_num$MahaD <- mahalanobis(df_sample_num,colMeans(df_sample_num),cov(df_sample_num))
```
```{r}
df_sample_num$Maha_P <- pchisq(df_sample_num$MahaD, df=(ncol(df_sample_num)-1), lower.tail=FALSE)
df_sample_num$outlier <- "No"
df_sample_num$outlier[df_sample_num$Maha_P < 0.001] <- "Yes"

#keep loan is bad for later
bad_loans <- df_sample_num$loan_is_bad
#remove outliers
df_sample_num <- df_sample_num %>%
  filter(outlier == "No") %>%
  select(-c("MahaD", "Maha_P", "outlier"))
bad_loans <- df_sample_num$loan_is_bad

df_sample_num <- df_sample_num %>%
  select(-c("loan_is_bad"))
```

Checking data again
```{r}
lowerCor(df_sample_num)
```
#Standardise data
```{r}
df_sample_num<-as.data.frame(scale(df_sample_num))
```
Testing if suitable for factor analysis
```{r}
KMO(df_sample_num)
cortest.bartlett(df_sample_num)
```
# Doing PCA
```{r}
pcModel<-principal(df_sample_num, 11, rotate="none")
print(pcModel)
```
```{r}
print.psych(pcModel, cut=0.3, sort=TRUE)
```
```{r}
plot(pcModel$values, type="b")
```
#Factor analysis with ML

```{r}
FA<-(fa(df_sample_num,4, n.obs=nrow(df_sample_num), rotate="none", fm="ml"))
print(FA, cut=0.3,sort="TRUE")
fa.diagram(FA)
```
#Factor analysis with Oblimin 
```{r}
FA_Obl<-(fa(df_sample_num,4, n.obs=nrow(df_sample_num), rotate="oblimin", fm="ml"))
print(FA_Obl, cut=0.3,sort="TRUE")
fa.diagram(FA_Obl)
```
#Factor analysis with Varimax 
```{r}
FA_Vm<-(fa(df_sample_num,4, n.obs=nrow(df_sample_num), rotate="Varimax", fm="ml"))
print(FA_Vm, cut=0.3,sort="TRUE")
fa.diagram(FA_Vm)
```
#PCA
```{r}
pc_model<-principal(df_sample_num,4, rotate="none")
print.psych(pc_model, cut=0.3, sort=TRUE)
plot(pc_model$values, type="b")

```
#PC Factor Analysis - Oblimin
```{r}
pc_model_obl <- principal(df_sample_num, 4, rotate="oblimin")
print.psych(pc_model_obl, cut=0.3, sort=TRUE)
```
#PC Factor Analysis - Quartimax
```{r}
pc_model_q <- principal(df_sample_num, 4, rotate="quartimax")
print.psych(pc_model_q, cut=0.3, sort=TRUE)

```
#Factor analysis
```{r}
#choosing oblimin rotation with 4 factors
pc_model_obl <- principal(df_sample_num, 4, rotate="oblimin",scores=TRUE)
fscores <- pc_model_obl$scores
```
```{r}
describe(fscores)
```
```{r}
headTail(fscores)
```
```{r}
FscoresMatrix<-cor(fscores)
print(FscoresMatrix)
```
```{r}
round(FscoresMatrix, 2)
```
```{r}
lowerCor(fscores)
```
#Cluster Analysis
Define linkage methods
```{r}
m <- c( "average", "single", "complete", "ward")
names(m) <- c( "average", "single", "complete", "ward")
```
```{r}
#Function to compute agglomerative coefficient
ac <- function(x) {
  agnes(df_sample_num, method = x)$ac
}
```
Calculate agglomerative coefficient for each clustering linkage method
```{r}
# agglomerative coefficient for all clustering methods
sapply(m, ac)
```
We should choose Ward’s minimum variance method as it produces the highest agglomerative coefficient
```{r}
gap_stat_h <- clusGap(df_sample_num, FUN = hcut, nstart = 25, K.max = 10, B = 50)
gap_stat_k <- clusGap(df_sample_num, FUN = kmeans, nstart = 25, K.max = 10, B = 50)
```
produce plot of clusters vs. gap statistic
```{r}
fviz_gap_stat(gap_stat_h)
fviz_gap_stat(gap_stat_k)
```
From the plot we can see that the gap statistic is high at  k = 10
Finding distance matrix
```{r}
distance_mat <- dist(df_sample_num, method = 'euclidean')
```
Fitting HC Model to dataset
```{r}
set.seed(240)  # Setting seed
Hierar_cl <- hclust(distance_mat, method = "ward")
Hierar_cl
```
Plotting dendrogram
```{r}
plot(Hierar_cl)
```
Cut tree by number of clusters
```{r}
fit <- cutree(Hierar_cl, k =10)
fit
```
Find observations in each cluster
```{r}
table(fit)
```
Append cluster labels to original data
```{r}
final_data <-cbind(df_sample_num, cluster = fit)
```
```{r}
head(final_data)
```
```{r}
df_clust <-as.data.frame(cbind(cluster = fit, loan_is_bad = bad_loans))
```
```{r}
#Summary stats
cluster_loans <- df_clust %>%
  group_by(cluster)%>%
  summarise(count_bad_loans = sum(loan_is_bad), total_in_cluster = n(), perc_bad = count_bad_loans/total_in_cluster)

cluster_loans
```



